PanGraph: scalable bacterial pan-genome graph construction

### 2022-11-23 main.tex
fix: algorithm figure
Update on Overleaf.

In recent years, efforts have focused on generalizing the \emph{pangenome} framework of microbial diversity to graphical models \cite{eizenga2020pangenome}.
At a high level, pangenome graphs generalize the reference sequence coordinate system conventionally used and encode genomes as directed paths through the graph consisting of node that represent sequence.
In multicellular eukaryotes, pangenome graphs usually have an overall linear structure in which structural diversity can be encoded as short range  excursions. 
In microbial genomics, in contrast, large scale rearrangements and inversions give rise to complex graph topologies. 
Here, we focus on pangenome graphs of closely related bacterial genomes.
\Richard{I feel the following is more confusing than helping.}
As such, pangenome graphs can be thought of as an analogous construct to the well-known assembly graph, in which whole genomes are individual ``reads" of the underlying microbial structure \cite{myers2005fragment}.

While easy to conceptualize, the construction of pangenome graphs has proven computationally challenging.
Colored generalizations of the de Bruijin graph-based assemblers have been successively used to build graphs from large sequence sets, although the underlying efficiency derives from a fixed k-mer size which prevents modelling long-range homology \cite{iqbal2012novo,muggli2017succinct}.
An orthogonal approach has been to formulate the inference of the pangenome graph as a multiple genome alignment.
However, current methods either scale poorly to large sets of genomes \cite{darling2010progressivemauve}, focus on comparisons across diverse sets of species across the tree of life at the cost of memory \cite{armstrong2020progressive}, or utilize a reference-guided approach by partitioning genomes first into annotated genes \cite{gautreau2020ppanggolin,colquhoun2021pandora}.

Here we present PanGraph, a Julia \cite{bezanson2017julia} library and command line interface, designed to efficiently align large sets of closely related genomes into a pangenome graph on personal computers.
The resulting graph both compresses the input sequence set and succinctly captures the population diversity at multiple scales: from nucleotide mutations and indels to structural polymorphisms driven by inversions, rearrangements, and gene gain/loss.
The underlying graph data structure can be exported into numerous formats for downstream analysis and visualization in software such as Bandage \cite{wick2015bandage}.
%The algorithm, along with extensive documentation, examples, and instructions for installation, are open source and can be found on GitHub.
%We anticipate PanGraph to be immediately useful in understanding microbial evolution in natural populations.

@@

\begin{figure}[htb]
    \includegraphics[width=.45\textwidth]{figs/algorithm.png}
    \caption{{\bf Overview of PanGraph algorithm}.
        (A) The alignment graph is constructed progressively by aligning graphs pairwise up a guide tree constructed from neighbor-joining the minimizer overlap between strains.
        (B) During pairwise alignment, \emph{pancontigs} (blue and green)
        are merged by identifying homologous intervals (shown in yellow).
        If the underlying alignments are viewed compatible, i.e. the energy is less than 0, the pancontigs are merged.
    }
    \label{fig:visualization}
\end{figure}

@@

In addition to being a useful operation, marginalization allows us to quantify the consistency and robustness of pangraph construction on real data, where the ground truth is unknown.
To verify that marginalized graphs are compatible with newly-built graphs, we built a pangenome graph for 50 randomly selected strains from the \textit{K.~pneumoniae} dataset. We then picked 50 random pairs of strains from the same set and built two graphs for each pair: the graph obtained by marginalizing the complete graph on the pair ( (i.e. marginalized graph in Fig.~\ref{fig:marginalization}A top), and the graph built directly from the pair (i.e. pairwise graph in Fig.~\ref{fig:marginalization}A top).
In order to compare both graphs, we compute the partition they generated on the genome of the isolates they include. Each genome is partitioned in pancontigs that are either shared on the pair, or private to one isolate (cf. Fig.~\ref{fig:marginalization}A bottom). We can classify the segments in the intersection of the two partitions in different categories, depending on whether the two partitions agree on whether the segment is shared or private. Moreover segments on which the two partitions agree are further sub-divided depending on whether they are private or shared on both partitions.\\
The compatibility between the two graphs requires segments on which the two partitions disagree to be few and short. Indeed, we verified that over the pairs we picked, these segments cover a very small fraction of the genome ($<1\%$) and have average size compatible with the default 100 bp precision threshold of \textit{PanGraph} (cf. Fig.~\ref{fig:marginalization}B-C). Conversely, segments on which the partitions agree have average size of several kbps. The fact that the two partitions almost completely agree cannot be explained simply by the fact that most of the genome is shared, since segments that are shared on both graphs cover on average only 85\% of the sequence. We confirmed these results using the fraction of shared k-mers, using an approach inspired by PopPUNK \cite{lees2019fast}. Namely, we approximate the fraction of homologous sequence between any two pairs using the fraction of shared k-mers ($k=21$) divided by $(1-d)^{k}$, where $d$ is the average pairwise divergence on core genes for the pair considered. This divisor corrects for k-mers on homologous sequence that are not shared due to mutations. The resulting distribution of shared sequence is in very good agreement with the fraction of segments that are shared on both graphs (cf. Fig.~\ref{fig:marginalization}B), suggesting that homologous sequences are correctly merged on the complete, marginalized and pairwise graphs.\\
We performed the same test on species from the other datasets, obtaining similar results on all species whose divergence is compatible with \textit{PanGraph} capabilities (cf. SI Section IV and SI Fig.~5).

@@

\section*{Acknowledgement}
We are grateful to Boris Shraiman for stimulating discussions.
This study was funded by the University of Basel and the NSF.
\Richard{Liam: do you need to add a funding statement?}

>>>>

In recent years, efforts have focused on generalizing the \emph{pangenome} framework of microbial diversity to graphical models \cite{eizenga2020pangenome}.
At a high level, pangenome graphs generalize the reference sequence coordinate system conventionally used and encode genomes as directed paths through the graph consisting of node that represent sequence.
In multicellular eukaryotes, pangenome graphs usually have an overall linear structure in which structural diversity can be encoded as short range  excursions.
In microbial genomics, in contrast, large scale rearrangements and inversions give rise to complex graph topologies.
Here, we focus on pangenome graphs of closely related bacterial genomes.

While easy to conceptualize, the construction of pangenome graphs has proven computationally challenging.
Colored generalizations of the de Bruijn graph-based assemblers have been successively used to build graphs from large sequence sets, although the underlying efficiency derives from a fixed k-mer size which prevents modelling long-range homology \cite{iqbal2012novo,muggli2017succinct}.
An orthogonal approach has been to formulate the inference of the pangenome graph as a multiple genome alignment.
However, current methods either scale poorly to large sets of genomes \cite{darling2010progressivemauve}, focus on comparisons across diverse sets of species across the tree of life at the cost of memory \cite{armstrong2020progressive}, or utilize a reference-guided approach by partitioning genomes first into annotated genes \cite{gautreau2020ppanggolin,colquhoun2021pandora}.

Here we present \emph{PanGraph}, a Julia \cite{bezanson2017julia} library and command line interface, designed to efficiently align large sets of closely related genomes into a pangenome graph on personal computers.
The resulting graph both compresses the input sequence set and succinctly captures the population diversity at multiple scales: from nucleotide mutations and indels to structural polymorphisms driven by inversions, rearrangements, and gene gain/loss.
The underlying graph data structure can be exported into numerous formats for downstream analysis and visualization in software such as Bandage \cite{wick2015bandage}.
%The algorithm, along with extensive documentation, examples, and instructions for installation, are open source and can be found on GitHub.
%We anticipate PanGraph to be immediately useful in understanding microbial evolution in natural populations.

@@

\begin{figure}[htb]
    \includegraphics[width=.45\textwidth]{figs/algorithm.pdf}
    \caption{{\bf Overview of PanGraph algorithm}.
        (A) The alignment graph is constructed progressively by aligning graphs pairwise up a guide tree constructed from neighbor-joining the minimizer overlap between strains.
        (B) During pairwise alignment, \emph{pancontigs} (blue and green)
        are merged by identifying homologous intervals (shown in yellow).
        If the underlying alignments are viewed compatible, i.e. the energy is less than 0, the pancontigs are merged.
    }
    \label{fig:visualization}
\end{figure}

@@

In addition to being a useful operation, marginalization allows us to quantify the consistency and robustness of pangraph construction on real data, where the ground truth is unknown.
To verify that marginalized graphs are compatible with newly-built graphs, we built a pangenome graph for 50 randomly selected strains from the \textit{K.~pneumoniae} dataset. We then picked 50 random pairs of strains from the same set and built two graphs for each pair: the graph obtained by marginalizing the complete graph on the pair (i.e. marginalized graph in Fig.~\ref{fig:marginalization}A top), and the graph built directly from the pair (i.e. pairwise graph in Fig.~\ref{fig:marginalization}A top).
In order to compare both graphs, we compute the partition they generated on the genome of the isolates they include. Each genome is partitioned in pancontigs that are either shared on the pair, or private to one isolate (cf. Fig.~\ref{fig:marginalization}A bottom). We can classify the segments in the intersection of the two partitions in different categories, depending on whether the two partitions agree on whether the segment is shared or private. Moreover segments on which the two partitions agree are further sub-divided depending on whether they are private or shared on both partitions.\\
The compatibility between the two graphs requires segments on which the two partitions disagree to be few and short. Indeed, we verified that over the pairs we picked, these segments cover a very small fraction of the genome ($<1\%$) and have average size compatible with the default 100 bp precision threshold of \textit{PanGraph} (cf. Fig.~\ref{fig:marginalization}B-C). Conversely, segments on which the partitions agree have average size of several kbps. The fact that the two partitions almost completely agree cannot be explained simply by the fact that most of the genome is shared, since segments that are shared on both graphs cover on average only 85\% of the sequence. We confirmed these results using the fraction of shared k-mers, using an approach inspired by PopPUNK \cite{lees2019fast}. Namely, we approximate the fraction of homologous sequence between any two pairs using the fraction of shared k-mers ($k=21$) divided by $(1-d)^{k}$, where $d$ is the average pairwise divergence on core genes for the pair considered. This divisor corrects for k-mers on homologous sequence that are not shared due to mutations. The resulting distribution of shared sequence is in very good agreement with the fraction of segments that are shared on both graphs (cf. Fig.~\ref{fig:marginalization}B), suggesting that homologous sequences are correctly merged on the complete, marginalized and pairwise graphs.\\
We performed the same test on species from the other datasets, obtaining similar results on all species whose divergence is compatible with \textit{PanGraph} capabilities (cf. SI Section IV and SI Fig.~5).

@@

\section*{Acknowledgement}
We are grateful to Boris Shraiman for stimulating discussions.
This study was funded by the University of Basel and the NSF.
L.P.S. is a Sir Henry Wellcome Postdoctoral Fellow funded by Wellcome (Grant 220422/Z/20/Z).
